---
title: "Reading NESSTAR survey data with nesstarR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading NESSTAR survey data with nesstarR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

mospi_data <- {
  env <- Sys.getenv("MOSPI_DATA", unset = "")
  if (nzchar(env)) {
    normalizePath(env, mustWork = FALSE)
  } else {
    # Sibling-repo default: nesstarR/ and mospiR/ sit next to each other
    normalizePath(file.path("..", "..", "mospiR", "data", "hces"),
                  mustWork = FALSE)
  }
}

nesstar_path <- file.path(
  mospi_data,
  "DDI-IND-MOSPI-NSSO-68Rnd-Sch2.0-July2011-June2012",
  "Nss68_1.0_Type2_new format",
  "survey0", "data",
  "nss68_consumer_expenditure_type2.Nesstar"
)

stata_zip <- normalizePath(
  file.path("..", "stata", "Nss68_STATA_2012.zip"),
  mustWork = FALSE
)
```

[NSS rounds 64, 66, and 68](https://microdata.gov.in/NADA/index.php/catalog/?page=1&sort_order=desc&ps=15) come as `.Nesstar` binaries inside `.rar` archives.
Download them with the sister package [mospiR](https://github.com/saketkc/mospiR), extract with
`unar`, and read with nesstarR. All examples below use NSS 68th Round Type 2
(Consumer Expenditure Survey, July 2011 - June 2012); the same steps apply to
NSS 64 and 66.

## Prerequisites

```{r install, eval=FALSE}
remotes::install_github("saketkc/mospiR")
remotes::install_github("saketkc/nesstarR")
```

```{r libs}
library(mospiR)
library(nesstarR)
```

## Download from the portal

`mospiR::download_dataset()` authenticates with the MoSPI NADA API and
downloads the full archive for one survey idno. The API key is read from
`MOSPI_KEY` (set it in `~/.Renviron`).

```{r download, eval=FALSE}
api_key  <- Sys.getenv("MOSPI_KEY")
data_dir <- file.path("data", "hces",
                      "DDI-IND-MOSPI-NSSO-68Rnd-Sch2.0-July2011-June2012")

download_dataset(
  "DDI-IND-MOSPI-NSSO-68Rnd-Sch2.0-July2011-June2012",
  data_dir,
  api_key
)
```

The download is a `.rar` archive, around 270 MB for NSS 68 T2.

## Extract the archive

`unar` (macOS/Linux) and `unrar` (Windows) both work. Either must be on your
`PATH`.

```{r extract, eval=FALSE}
rar_file <- file.path(data_dir, "Nss68_1.0_Type2_new format.rar")
system2("unar", c("-o", dirname(rar_file), shQuote(rar_file)))
```

Extraction produces a folder containing:

- `survey0/data/nss68_consumer_expenditure_type2.Nesstar` (the binary data)
- `ddi.xml` (variable metadata in DDI format, not required by nesstarR)
- `Readme.pdf` / supporting documents

## Parse the Nesstar file

`nesstar_parse()` reads the entire file into memory and decodes the binary
header, resource index, dataset descriptors, and variable directories. No row
data is loaded at this stage.

```{r parse}
nb <- nesstar_parse(nesstar_path)
nb
```

## Dataset structure

A single `.Nesstar` container holds multiple datasets, one per schedule
block (household, person, food, non-food, …).

```{r datasets}
nesstar_datasets(nb)
```

The dataset with the largest row count is the food block: one row per
household × item combination (~4-5 million rows for NSS 68 T2).

```{r food_ds}
ds_tbl <- nesstar_datasets(nb)
food_ds <- ds_tbl$dataset_number[which.max(ds_tbl$row_count)]
cat("Food block dataset number:", food_ds, "\n")
cat("Rows:", ds_tbl$row_count[ds_tbl$dataset_number == food_ds], "\n")
```

## Variable listing

```{r variables}
vars <- nesstar_variables(nb, dataset_number = food_ds)
vars[, c("name", "mode_code", "value_format_code", "width_value")]
```

## Embedded labels and category codes

NESSTAR files from NSS 64 onwards embed Huffman-compressed XML blocks with
variable labels and category codes.

```{r metadata}
meta <- nesstar_metadata(nb)
if (!is.null(meta)) {
  ds_meta <- meta$datasets[[which(
    vapply(meta$datasets, `[[`, integer(1), "dataset_number") == food_ds
  )]]
  cat("File name stored in container:", ds_meta$file_name, "\n\n")

  for (v in head(ds_meta$variables, 6)) {
    cat(sprintf("%-20s  label: %s\n", v$name, v$label))
    if (length(v$categories) > 0) {
      for (cat_entry in head(v$categories, 4)) {
        cat(sprintf("  %s = %s\n", cat_entry$value, cat_entry$label))
      }
    }
  }
} else {
  message("Metadata not available for this file (older NSS format).")
}
```


## Read a dataset

`nesstar_read_dataset()` decodes all columns for the requested dataset.
For the food block this is the slow step (~10-30 s on a laptop): 4–5 million
rows × ~20 columns, decoded from the packed binary format.

Pass `columns` to load only what you need:

```{r read}
food <- nesstar_read_dataset(
  nb,
  dataset_number = food_ds,
  columns = c("HHID", "State_code", "District_Code", "Sector",
              "Combined_Multiplier", "Item_Code", "Total_Consumption_Value")
)
cat("Rows:", nrow(food), "| Columns:", ncol(food), "\n")
head(food)
```

## Quick check: cereal spending by sector

Weighted mean monthly household expenditure on cereals (item codes 101–140),
rural vs. urban. The `Multiplier` column is the survey weight.

```{r cereals}
cereals    <- food[as.integer(food$Item_Code) >= 101 &
                   as.integer(food$Item_Code) <= 140, ]
hh_cereals <- aggregate(Total_Consumption_Value ~ HHID + Sector + Combined_Multiplier,
                        data = cereals, FUN = sum)

rural <- hh_cereals[hh_cereals$Sector == "1", ]
urban <- hh_cereals[hh_cereals$Sector == "2", ]

cat(sprintf(
  "Weighted mean cereal expenditure (Rs/month):\n  Rural: %.0f\n  Urban: %.0f\n",
  weighted.mean(rural$Total_Consumption_Value, rural$Combined_Multiplier, na.rm = TRUE),
  weighted.mean(urban$Total_Consumption_Value, urban$Combined_Multiplier, na.rm = TRUE)
))
```

## Export to csv

`nesstar_export()` writes one CSV (or `.csv.gz`) per dataset to a directory.
It processes 50,000 rows per chunk by default, so even the 4.65 GB T1 file
does not require loading all rows at once.

```{r export}
output_dir <- file.path(tempdir(), "nss68t2")
nesstar_export(nb, output_dir = output_dir, compress = FALSE)
list.files(output_dir)
```

Pass `compress = TRUE` (the default) for gzip-compressed output, which
typically halves file size with negligible extra time.

```{r export_subset, eval=FALSE}
nesstar_export(nb,
               output_dir = output_dir,
               datasets   = food_ds,
               compress   = TRUE,
               chunk_size = 100000L)
```

## Verify against Stata reference files

For this, I took help of a colleague to install Nesstar.exe and asked them to convert the binary to stata file. 
The code below checks that nesstarR's CSV output matches it.

### Unzip the Stata archive

```{r stata_unzip}
stata_dir <- file.path(tempdir(), "stata_68t2")
unzip(stata_zip, exdir = stata_dir)
dta_files <- sort(list.files(stata_dir, pattern = "\\.dta$",
                              full.names = TRUE, recursive = TRUE))
cat(length(dta_files), "Stata files found\n")
```

### Match datasets by column set

The Stata archive labels each file "Level N". nesstarR names output files
`_dsNN`. We read only column schemas from the Stata files (`n_max = 0` loads
the header without row data) and match each level to a CSV by its normalised
column name set, which is unique across all 11 datasets.

```{r stata_match}
library(haven)

# n_max = 0: read header only, no row data loaded
stata_meta <- lapply(dta_files, function(f) {
  df <- read_dta(f, n_max = 0L)
  list(
    path  = f,
    level = as.integer(sub(".*[Ll]evel[_\\s-]*(\\d+).*", "\\1",
                           basename(f), perl = TRUE)),
    ncol  = ncol(df),
    cols  = names(df)
  )
})

csv_files <- sort(list.files(output_dir, pattern = "_ds\\d+\\.csv$",
                              full.names = TRUE))
csv_meta <- lapply(csv_files, function(f) {
  hdr    <- names(read.csv(f, nrows = 0L, check.names = FALSE))
  ds_num <- as.integer(sub(".*_ds(\\d+)\\.csv$", "\\1", basename(f)))
  list(path = f, ds_num = ds_num, cols = hdr)
})

# STATA converts '.' in column names to '_'; normalise before comparing
norm_names <- function(x) toupper(gsub("\\.", "_", trimws(x)))

for (sm in stata_meta) {
  sn <- norm_names(sm$cols)
  cm <- Filter(function(x) setequal(norm_names(x$cols), sn), csv_meta)
  if (length(cm) == 0L) {
    cat(sprintf("Level %2d  cols=%d  -> NO MATCH\n", sm$level, sm$ncol))
    next
  }
  cat(sprintf("Level %2d -> ds%02d  cols=%d  cols_match=TRUE\n",
              sm$level, cm[[1L]]$ds_num, sm$ncol))
}
```

Expected output: every Stata level maps to exactly one nesstarR CSV dataset.

### Spot-check numeric values

Compare the first 500 rows of each matched dataset cell by cell. Any
difference beyond floating-point epsilon (`1e-9`) is flagged.

```{r stata_values}
SPOT_ROWS <- 500L

for (sm in stata_meta) {
  sn <- norm_names(sm$cols)
  cm <- Filter(function(x) setequal(norm_names(x$cols), sn), csv_meta)
  if (length(cm) == 0L) next
  cm <- cm[[1L]]

  s_df <- read_dta(sm$path, n_max = SPOT_ROWS)
  c_df <- read.csv(cm$path, nrows = SPOT_ROWS, check.names = FALSE)

  s_norm <- norm_names(names(s_df))
  c_norm <- norm_names(names(c_df))
  common <- intersect(s_norm, c_norm)

  mismatches <- 0L
  for (col in common) {
    sv <- suppressWarnings(as.numeric(as.character(
            s_df[[which(s_norm == col)[1L]]])))
    cv <- suppressWarnings(as.numeric(as.character(
            c_df[[which(c_norm == col)[1L]]])))
    if (!all(is.na(sv)) && !all(is.na(cv))) {
      mismatches <- mismatches +
        sum(abs(sv - cv) > 1e-9, na.rm = TRUE) +
        sum(is.na(sv) & !is.na(cv)) +
        sum(!is.na(sv) & is.na(cv))
    } else {
      mismatches <- mismatches +
        sum(trimws(as.character(sv)) != trimws(as.character(cv)), na.rm = TRUE)
    }
  }

  cat(sprintf(
    "Level %2d / ds%02d: %d cols x %d rows - %s\n",
    sm$level, cm$ds_num, length(common), SPOT_ROWS,
    if (mismatches == 0L) "all values match" else paste(mismatches, "MISMATCHES")
  ))
}
```

