---
title: "Reading NESSTAR survey data with nesstarR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading NESSTAR survey data with nesstarR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

mospi_data <- {
  env <- Sys.getenv("MOSPI_DATA", unset = "")
  if (nzchar(env)) {
    normalizePath(env, mustWork = FALSE)
  } else {
    # Sibling-repo default: nesstarR/ and mospiR/ sit next to each other
    normalizePath(file.path("..", "..", "mospiR", "data", "hces"),
      mustWork = FALSE
    )
  }
}

nesstar_path <- file.path(
  mospi_data,
  "DDI-IND-MOSPI-NSSO-68Rnd-Sch2.0-July2011-June2012",
  "Nss68_1.0_Type2_new format",
  "survey0", "data",
  "nss68_consumer_expenditure_type2.Nesstar"
)

stata_zip <- normalizePath(
  file.path("..", "stata", "Nss68_STATA_2012.zip"),
  mustWork = FALSE
)
```

[NSS rounds 64, 66, and 68](https://microdata.gov.in/NADA/index.php/catalog/?page=1&sort_order=desc&ps=15) come as `.Nesstar` binaries inside `.rar` archives.
Download them with the sister package [mospiR](https://github.com/saketkc/mospiR), extract with
`unar`, and read with nesstarR. All examples below use NSS 68th Round Type 2
(Consumer Expenditure Survey, July 2011 - June 2012); the same steps apply to
NSS 64 and 66.

## Prerequisites

```{r install, eval=FALSE}
remotes::install_github("saketkc/mospiR")
remotes::install_github("saketkc/nesstarR")
```

```{r libs}
library(mospiR)
library(nesstarR)
```

## Download from the portal

`mospiR::download_dataset()` authenticates with the MoSPI NADA API and
downloads the full archive for one survey idno. The API key is read from
`MOSPI_KEY` (set it in `~/.Renviron`).

```{r download, eval=FALSE}
api_key <- Sys.getenv("MOSPI_KEY")
data_dir <- file.path(
  "data", "hces",
  "DDI-IND-MOSPI-NSSO-68Rnd-Sch2.0-July2011-June2012"
)

download_dataset(
  "DDI-IND-MOSPI-NSSO-68Rnd-Sch2.0-July2011-June2012",
  data_dir,
  api_key
)
```

The download is a `.rar` archive, around 270 MB for NSS 68 T2.

## Extract the archive

`unar` (macOS/Linux) and `unrar` (Windows) both work. Either must be on your
`PATH`.

```{r extract, eval=FALSE}
rar_file <- file.path(data_dir, "Nss68_1.0_Type2_new format.rar")
system2("unar", c("-o", dirname(rar_file), shQuote(rar_file)))
```

Extraction produces a folder containing:

- `survey0/data/nss68_consumer_expenditure_type2.Nesstar` (the binary data)
- `ddi.xml` (variable metadata in DDI format, not required by nesstarR)
- `Readme.pdf` / supporting documents

## Parse the Nesstar file

`nesstar_parse()` reads the entire file into memory and decodes the binary
header, resource index, dataset descriptors, and variable directories. No row
data is loaded at this stage.

```{r parse}
nb <- nesstar_parse(nesstar_path)
nb
```

## Dataset structure

A single `.Nesstar` container holds multiple datasets, one per schedule
block (household, person, food, non-food, …).

```{r datasets}
nesstar_datasets(nb)
```

The dataset with the largest row count is the food block: one row per
household × item combination (~4-5 million rows for NSS 68 T2).

```{r food_ds}
ds_tbl <- nesstar_datasets(nb)
food_ds <- ds_tbl$dataset_number[which.max(ds_tbl$row_count)]
cat("Food block dataset number:", food_ds, "\n")
cat("Rows:", ds_tbl$row_count[ds_tbl$dataset_number == food_ds], "\n")
```

## Variable listing

```{r variables}
vars <- nesstar_variables(nb, dataset_number = food_ds)
vars[, c("name", "mode_code", "value_format_code", "width_value")]
```

## Embedded labels and category codes

NESSTAR files from NSS 64 onwards embed Huffman-compressed XML blocks with
variable labels and category codes.

```{r metadata}
meta <- nesstar_metadata(nb)
if (!is.null(meta)) {
  ds_meta <- meta$datasets[[which(
    vapply(meta$datasets, `[[`, integer(1), "dataset_number") == food_ds
  )]]
  cat("File name stored in container:", ds_meta$file_name, "\n\n")

  for (v in head(ds_meta$variables, 6)) {
    cat(sprintf("%-20s  label: %s\n", v$name, v$label))
    if (length(v$categories) > 0) {
      for (cat_entry in head(v$categories, 4)) {
        cat(sprintf("  %s = %s\n", cat_entry$value, cat_entry$label))
      }
    }
  }
} else {
  message("Metadata not available for this file (older NSS format).")
}
```


## Read a dataset

`nesstar_read_dataset()` decodes all columns for the requested dataset.
For the food block this is the slow step (~10-30 s on a laptop): 4–5 million
rows × ~20 columns, decoded from the packed binary format.

Pass `columns` to load only what you need:

```{r read}
food <- nesstar_read_dataset(
  nb,
  dataset_number = food_ds,
  columns = c(
    "HHID", "State_code", "District_Code", "Sector",
    "Combined_Multiplier", "Item_Code", "Total_Consumption_Value"
  )
)
cat("Rows:", nrow(food), "| Columns:", ncol(food), "\n")
head(food)
```

## Quick check: cereal spending by sector

Weighted mean monthly household expenditure on cereals (item codes 101–140),
rural vs. urban. The `Multiplier` column is the survey weight.

```{r cereals}
cereals <- food[as.integer(food$Item_Code) >= 101 &
  as.integer(food$Item_Code) <= 140, ]
hh_cereals <- aggregate(Total_Consumption_Value ~ HHID + Sector + Combined_Multiplier,
  data = cereals, FUN = sum
)

rural <- hh_cereals[hh_cereals$Sector == "1", ]
urban <- hh_cereals[hh_cereals$Sector == "2", ]

cat(sprintf(
  "Weighted mean cereal expenditure (Rs/month):\n  Rural: %.0f\n  Urban: %.0f\n",
  weighted.mean(rural$Total_Consumption_Value, rural$Combined_Multiplier, na.rm = TRUE),
  weighted.mean(urban$Total_Consumption_Value, urban$Combined_Multiplier, na.rm = TRUE)
))
```

## Export to csv

`nesstar_export()` writes one CSV (or `.csv.gz`) per dataset to a directory.
It processes 50,000 rows per chunk by default, so even the 4.65 GB T1 file
does not require loading all rows at once.

```{r export}
# Write CSVs alongside the .Nesstar file so they survive across R sessions
# and avoid filling /tmp on machines with limited temporary storage.
output_dir <- dirname(nesstar_path)
nesstar_export(nb, output_dir = output_dir, compress = FALSE)
list.files(output_dir, pattern = "\\.csv$")
```

Pass `compress = TRUE` (the default) for gzip-compressed output, which
typically halves file size with negligible extra time.

```{r export_subset, eval=FALSE}
nesstar_export(nb,
  output_dir = output_dir,
  datasets   = food_ds,
  compress   = TRUE,
  chunk_size = 100000L
)
```

## Verify against Stata reference files

The original NESSTAR Publisher software was used to export each dataset to Stata `.dta` format.
The code below loads those exports and checks that nesstarR's CSV output matches them row-for-row.

### Unzip the Stata archive

```{r stata_unzip}
# Extract alongside the .Nesstar output to avoid exhausting /tmp on
# space-constrained machines (the full STATA archive expands to ~1.4 GB).
stata_dir <- file.path(output_dir, "stata_68t2")
unzip(stata_zip, exdir = stata_dir)
dta_files <- sort(list.files(stata_dir,
  pattern = "\\.dta$",
  full.names = TRUE, recursive = TRUE
))
cat(length(dta_files), "Stata files found\n")
```

### Build the dataset correspondence table

The NESSTAR container stores a file name for every dataset in its embedded
metadata (`<FileName>` XML field decoded by `nesstar_metadata()`). For NSS
files this is something like `Level_1.sav`. We extract the level number from
that string and join to the STATA export filenames — a structural match that
is independent of any data property. If the metadata names do not carry level
numbers we fall back to matching on unique `(nrow, ncol)` pairs.

```{r stata_match}
library(haven)

# Extract the integer N from "Level_N" / "Level N" / "LevelN" etc.
level_num <- function(s) {
  if (is.na(s) || !nzchar(s)) {
    return(NA_integer_)
  }
  m <- regexpr("[Ll]evel[[:space:]_-]*(\\d+)", s, perl = TRUE)
  if (m[[1L]] == -1L) {
    return(NA_integer_)
  }
  cs <- attr(m, "capture.start")[[1L]]
  cl <- attr(m, "capture.length")[[1L]]
  as.integer(substr(s, cs, cs + cl - 1L))
}

# If the container has embedded XML metadata (NSS 64+), each dataset carries a
# file name such as "Level_1.sav". We extract the level number from that.
# Older containers (e.g. some NSS 68 files) return NULL; in that case every
# ds$level stays NA and we rely entirely on the (nrow, ncol) fallback below.
meta <- nesstar_metadata(nb)
nesstar_ds <- if (!is.null(meta)) {
  do.call(rbind, lapply(meta$datasets, function(dm) {
    data.frame(
      ds_num = dm$dataset_number,
      file_name = dm$file_name,
      level = level_num(dm$file_name),
      stringsAsFactors = FALSE
    )
  }))
} else {
  cat("No embedded XML metadata — falling back to (nrow, ncol) matching.\n")
  data.frame(
    ds_num = vapply(nb$datasets, `[[`, integer(1), "dataset_number"),
    file_name = NA_character_,
    level = NA_integer_,
    stringsAsFactors = FALSE
  )
}
if (!is.null(meta)) {
  print(nesstar_ds[, c("ds_num", "file_name", "level")])
}

# Read header only (n_max=0) to get column names; read row count from binary.
dta_nobs <- function(path) {
  raw <- readBin(path, "raw", n = 512L)
  if (length(raw) < 10L) {
    return(NA_integer_)
  }
  ver <- as.integer(raw[1L])
  if (ver %in% c(102L, 108L, 110L, 113L, 114L, 115L)) {
    bo <- if (as.integer(raw[2L]) == 1L) "big" else "little"
    return(readBin(raw[7L:10L], "integer", n = 1L, size = 4L, endian = bo))
  }
  lt <- as.raw(0x3c)
  Nb <- as.raw(0x4e)
  for (i in seq_len(length(raw) - 4L)) {
    if (raw[i] == lt && raw[i + 1L] == Nb && raw[i + 2L] == as.raw(0x3e)) {
      return(readBin(raw[(i + 3L):(i + 6L)], "integer",
        n = 1L, size = 4L,
        endian = "little"
      ))
    }
  }
  NA_integer_
}

stata_ds <- do.call(rbind, lapply(dta_files, function(f) {
  df <- read_dta(f, n_max = 0L)
  data.frame(
    stata_path = f,
    level = level_num(basename(f)),
    nrow = dta_nobs(f),
    ncol = ncol(df),
    stringsAsFactors = FALSE
  )
}))

csv_files <- sort(list.files(output_dir,
  pattern = "_ds\\d+\\.csv$",
  full.names = TRUE
))
csv_meta <- lapply(csv_files, function(f) {
  hdr <- names(read.csv(f, nrows = 0L, check.names = FALSE))
  wc_out <- system(paste("wc -l", shQuote(f)), intern = TRUE)
  nrows <- as.integer(sub("^\\s*(\\d+).*", "\\1", wc_out)) - 1L
  ds_num <- as.integer(sub(".*_ds(\\d+)\\.csv$", "\\1", basename(f)))
  list(ds_num = ds_num, path = f, nrow = nrows, ncol = length(hdr))
})

match_tbl <- data.frame(
  level = integer(), ds_num = integer(),
  csv_path = character(), stata_path = character(), method = character(),
  stringsAsFactors = FALSE
)

for (i in seq_len(nrow(stata_ds))) {
  sd <- stata_ds[i, ]

  # Primary: metadata level number
  nm <- nesstar_ds[!is.na(nesstar_ds$level) & nesstar_ds$level == sd$level, ]
  if (nrow(nm) == 1L) {
    csv_f <- Filter(function(cm) cm$ds_num == nm$ds_num, csv_meta)
    if (length(csv_f) == 1L) {
      match_tbl <- rbind(match_tbl, data.frame(
        level = sd$level, ds_num = nm$ds_num,
        csv_path = csv_f[[1L]]$path, stata_path = sd$stata_path,
        method = "metadata", stringsAsFactors = FALSE
      ))
      next
    }
  }

  # Fallback: unique (nrow, ncol) among not-yet-matched NESSTAR datasets
  taken <- match_tbl$ds_num
  cands <- Filter(function(cm) {
    !cm$ds_num %in% taken &&
      cm$nrow == sd$nrow
  }, csv_meta)
  if (length(cands) > 1L) {
    cands <- Filter(function(cm) cm$ncol == sd$ncol, cands)
  }
  if (length(cands) == 1L) {
    match_tbl <- rbind(match_tbl, data.frame(
      level = sd$level, ds_num = cands[[1L]]$ds_num,
      csv_path = cands[[1L]]$path, stata_path = sd$stata_path,
      method = "row+col", stringsAsFactors = FALSE
    ))
  } else {
    cat(sprintf(
      "Level %2d: COULD NOT MATCH (nrow=%d, ncol=%d)\n",
      sd$level, sd$nrow, sd$ncol
    ))
  }
}

match_tbl <- match_tbl[order(match_tbl$level), ]
cat(sprintf("\n%d / %d STATA levels matched:\n", nrow(match_tbl), nrow(stata_ds)))
for (i in seq_len(nrow(match_tbl))) {
  r <- match_tbl[i, ]
  cat(sprintf(
    "  Level %2d -> ds%02d  [%-8s]  %s\n",
    r$level, r$ds_num, r$method, basename(r$csv_path)
  ))
}
```

### Full dataset verification

Load every matched pair completely and compare all rows and all columns.
STATA converts `.` to `_` in variable names; `norm_col()` normalises both
sides. `haven::zap_labels()` strips value-label attributes so numeric
columns compare as plain doubles. Each dataset is freed from memory after
comparison to limit peak usage.

```{r stata_verify}
# Normalise column names: upper-case, any run of . or spaces -> _
norm_col <- function(x) toupper(gsub("[. ]+", "_", trimws(x)))

# Strip haven label / factor wrappers; return comparable base R vector
strip_stata <- function(x) {
  if (inherits(x, c("haven_labelled", "haven_labelled_spss"))) {
    return(as.numeric(unclass(x)))
  }
  if (is.factor(x)) {
    return(as.character(x))
  }
  x
}

pass_all <- TRUE

for (i in seq_len(nrow(match_tbl))) {
  r <- match_tbl[i, ]

  s_df <- tryCatch(
    as.data.frame(zap_labels(read_dta(r$stata_path))),
    error = function(e) {
      cat(sprintf(
        "Level %2d / ds%02d: SKIP  (STATA read error: %s)\n",
        r$level, r$ds_num, conditionMessage(e)
      ))
      NULL
    }
  )
  if (is.null(s_df)) {
    pass_all <- FALSE
    next
  }
  c_df <- read.csv(r$csv_path, check.names = FALSE)

  names(s_df) <- norm_col(names(s_df))
  names(c_df) <- norm_col(names(c_df))

  if (nrow(s_df) != nrow(c_df)) {
    cat(sprintf(
      "Level %2d / ds%02d: FAIL  row count STATA=%d  CSV=%d\n",
      r$level, r$ds_num, nrow(s_df), nrow(c_df)
    ))
    pass_all <- FALSE
    rm(s_df, c_df)
    gc(verbose = FALSE)
    next
  }

  only_s <- setdiff(names(s_df), names(c_df))
  only_c <- setdiff(names(c_df), names(s_df))
  common <- intersect(names(s_df), names(c_df))

  # Per-column comparison across ALL rows.
  # Many survey code columns (Sch_no, Level, …) are stored as zero-padded
  # strings in the NESSTAR binary and are preserved as such by nesstarR, but
  # read.csv() strips leading zeros when parsing quoted numeric-looking fields
  # (e.g. "010" → 10).  STATA also carries zero-padded display formats.  Both
  # representations encode the same integer value, so we compare numerically
  # whenever every non-NA cell on both sides is numeric-convertible.
  col_fail <- character(0)
  for (col in common) {
    sv <- strip_stata(s_df[[col]])
    cv <- c_df[[col]]
    sv_c <- trimws(as.character(sv))
    cv_c <- trimws(as.character(cv))
    sv_n <- suppressWarnings(as.numeric(sv_c))
    cv_n <- suppressWarnings(as.numeric(cv_c))
    # Extra NAs introduced by coercion (not originally NA or empty)?
    sv_was_na <- is.na(sv_c) | sv_c == ""
    cv_was_na <- is.na(cv_c) | cv_c == ""
    if (sum(is.na(sv_n) & !sv_was_na) == 0L &&
      sum(is.na(cv_n) & !cv_was_na) == 0L) {
      # Both sides fully numeric-representable: compare as numbers
      n_diff <- sum(abs(sv_n - cv_n) > 1e-9, na.rm = TRUE) +
        sum(xor(is.na(sv_n), is.na(cv_n)))
    } else {
      # At least one side has genuine string values: compare as character
      n_diff <- sum(sv_c != cv_c, na.rm = TRUE) +
        sum(xor(is.na(sv_c), is.na(cv_c)))
    }
    if (n_diff > 0L) col_fail <- c(col_fail, sprintf("%s(%d)", col, n_diff))
  }

  if (length(col_fail) == 0L) {
    cat(sprintf(
      "Level %2d / ds%02d: PASS  %7d rows x %d cols\n",
      r$level, r$ds_num, nrow(s_df), length(common)
    ))
  } else {
    pass_all <- FALSE
    cat(sprintf(
      "Level %2d / ds%02d: FAIL  %7d rows x %d cols | diffs: %s\n",
      r$level, r$ds_num, nrow(s_df), length(common),
      paste(col_fail, collapse = ", ")
    ))
  }
  if (length(only_s) > 0L) {
    cat(sprintf("         only in STATA: %s\n", paste(only_s, collapse = ", ")))
  }
  if (length(only_c) > 0L) {
    cat(sprintf("         only in CSV:   %s\n", paste(only_c, collapse = ", ")))
  }

  rm(s_df, c_df)
  gc(verbose = FALSE)
}

if (pass_all) {
  cat("\n==> ALL LEVELS PASS: nesstarR output is identical to the STATA reference export.\n")
} else {
  cat("\n==> Failures detected — see details above.\n")
}
```

